Supplementary material to: Drug efficacy prediction in cell lines

### 2019-01-08 docs/supplement.Rnw
rewrote simulation models in supplement.Rnw

% R packages
<<echo=FALSE>>=
library(cambridge)
@

<<figures, cache=FALSE, echo=FALSE>>=
read_chunk("../code/figures.R")
@

@@

\begin{document}
	\maketitle

@@

  \section{Model without tissue effects}\label{sec:notissuemodel}
  In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$. The model is given by:
  \begin{align*}
  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
  \gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda), \\
  \sigma_d^2 & \sim 1/\sigma_d^3.
  \end{align*}
  The corresponding variational Bayes posterior is given by:
  \begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
  \end{align*}
  We update the variational parameters until convergence:
  \begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} + \\
  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, a_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg],
  \end{align*}
  with $a_d^{(h)}$ as in the full model in the MD. Equivalently, the empirical Bayes updates for $\bm{\alpha}$ and $\lambda$ are the same as in the full model in the MD.

@@

  Its variational posterior is given by:
  \begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{p}{2} + a_{\text{class}(d)}, b^*_{\text{class}(d)}\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{n+p}{2} + 0.001, d^*_{\text{class}(d)} \),
  \end{align*}
  with variational parameter updates:
  \begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = \frac{n + p - 1.998}{2(d^*_{\text{class}(d)})^{(h)}}\(\X \tr \X + \frac{p + 2a_{\text{class}(d)} - 2}{2(b^*_{\text{class}(d)})^{(h)}} \I_p\)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = \(\X \tr \X + \frac{p + 2a_{\text{class}(d)} - 2}{2(b^*_{\text{class}(d)})^{(h)}} \I_p\)^{-1} \X \tr \y_d, \\
  b^*_{\text{class}(d)} & = b_{\text{class}(d)} + \frac{(d^*_{\text{class}(d)})^{(h)}}{n + p -1.998} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}],\\
  d^*_{\text{class}(d)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\].
  \end{align*}

>>>>

@@

\begin{document}

% load packages and set options
<<include=FALSE>>=
library(cambridge)
opts_knit$set(base.dir="../figs")
opts_chunk$set(fig.align='center', echo=FALSE, fig.path="../figs/")
@

% read figure code
<<figures, include=FALSE>>=
read_chunk("../code/figures.R")
@

	\maketitle

@@

	\section{Simulations}
	\subsection{Models}
	In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$. The model is given by:
	\begin{align*}
  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim p(\beta_{jd}), \\
  \gamma_d^2 & \sim p(\gamma_d^2), \\
  \sigma_d^2 & \sim 1/\sigma_d^3.
  \end{align*}
  We use a conjugate and non-conjugate model for $\beta_{jd}$:
	\begin{align*}
  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2),
  \end{align*}
	respectively. For the prior variances we use an inverse Gaussian and inverse Gamma model
	\begin{align*}
  \gamma_d^2 \sim \mathcal{IG}(\theta_d, \lambda_d) = \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
  \gamma_d^2 \sim \Gamma^{-1}(\alpha_d/2, \lambda_d/2) = \mathcal{GIG}(-\alpha_d/2, 0, \lambda_d),
  \end{align*}
	respectively. We have rescaled the inverse Gamma parameters to make them more comparable in the following. The two distributions coincide if $\theta_d \to \infty$ and $\alpha_d = 1$.

	\subsection{Estimation}
	The variational distributions are:
	\begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\alpha_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\),
  \end{align*}
	where $\alpha_d=1$ if we use the inverse Gaussian prior and $\theta_d \to \infty$ if we use the inverse Gamma prior. The parameters contain cyclic dependencies and are, in the conjugate setting updated by:
	\begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = a_d^{(h)} \bmu_d \tr \bmu_d + \lambda_d, \\ 
  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + b_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
  \end{align*}
	Likewise in the non-conjugate setting:
	\begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = \bmu_d \tr \bmu_d + \lambda_d, \\ 
  \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \].
  \end{align*}
  We have $a_d^{(h)}=\E_{Q^{(h)}}(\sigma_d^{-2})=(n+p+1)/(2 \zeta_d^{(h)})$ and
  $$
  b_d^{(h)} = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \alpha_d -2}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\alpha_d}{2}} \( \sqrt{\lambda \delta_d^{(h)}}/\theta_d \)} + \frac{p+\alpha_d^{(h)}}{\delta_d^{(h)}}.\\
  $$
  This simplifies to $b_d^{(h)}=(p+\alpha_d)/\delta_d^{(h)}$ in the inverse Gamma model.

  \subsection{Efficient computation}
  Let the SVD of $\X$ be $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$, $v_i$ be the singular values, and $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting. Then:
  \begin{align*}
  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
  \end{align*}
  The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm. The rest of the calculations involve just $n$ scalar multiplications and/or additions.

  \section{Model without tissue effects}\label{sec:notissuemodel}
  In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$. The model is given by:
  \begin{align*}
  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
  \gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda), \\
  \sigma_d^2 & \sim 1/\sigma_d^3.
  \end{align*}
  The corresponding variational Bayes posterior is given by:
  \begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
  \end{align*}
  We update the variational parameters until convergence:
  \begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ 
  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + a_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + a_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg],
  \end{align*}
  with $a_d^{(h)}$ as in the full model in the MD. Equivalently, the empirical Bayes updates for $\bm{\alpha}$ and $\lambda$ are the same as in the full model in the MD.

@@

  Its variational posterior is given by:
  \begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{p}{2} + a_{\text{class}(d)}, b^*_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \( \frac{n+p}{2} + 0.001, d^*_d \),
  \end{align*}
  with variational parameter updates:
  \begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = (c_d^{(h)})^{-1} \(\X \tr \X + (e_d^{(h)})^{-1} \I_p\)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = \(\X \tr \X + (e_d^{(h)})^{-1} \I_p\)^{-1} \X \tr \y_d, \\
  (b^*_d)^{(h+1)} & = b_{\text{class}(d)} + \frac{c_d^{(h)}}{2} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}],\\
  (d^*_d)^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\] + \\
  & \frac{e_d^{(h)}}{2} \trace ( \bm{\Sigma}_d^{(h+1)}) + \frac{e_d^{(h)}}{2} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)},
  \end{align*}
  where
  $$
  c_d^{(h)} = \frac{2(d_d^*)^{(h)}}{n + p -1.998} \text{ and } e_d^{(h)} = \frac{2(b_d^*)^{(h)}}{p + 2a_{\text{class}(d)} - 2}.
  $$

  \begin{align*}
  \text{ELBO}_d & (Q^{(l)}) \propto -\(\frac{n+p}{2} + 0.001 \) \log \[ (d^*_d)^{(l)}\] + \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \[ (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} + \trace (\bm{\Sigma}_d^{(l)}) \] \frac{c_d^{(l)} e_d^{(l)}}{2} \\
  & - \[ \bm{y}_d \tr \bm{y}_d - \bm{y}_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace (\X \tr \X \bm{\Sigma}_d^{(l)})\] \frac{c_d^{(l)}}{2} \\
  & + a_{\text{class}(d)}^{(l)} + \log \[\frac{\Gamma (a_{\text{class}(d)}^{(l)} + p/2)}{\Gamma (a_{\text{class}(d)}^{(l)})}\] - \( a_{\text{class}(d)}^{(l)} +\frac{p}{2} \) \log \[(b_d^*)^{(l)}\]  \\ 
  & + a_{\text{class}(d)}^{(l)} \log b_{\text{class}(d)}^{(l)} - b_{\text{class}(d)}^{(l)} \frac{a_{\text{class}(d)}^{(l)} + p/2}{(b_d^*)^{(h)}}.
  \end{align*}
